Classifying galaxy spectra at $0.5<z<1$: an unsupervised approach

### 2017-08-11 sections0.08_for_editor/intro0.08.tex
re-submitted version

Several attempts have been made to create a complete set of templates for categorizing the spectral type of galaxies using observations of nearby galaxies~(e.g.~\citealt{Kinney93}, \citealt[][hereafter \citetalias{Kinney96}]{Kinney96}, \citealt{Bershady00}, \citealt{Mannucci01}). 
Based on their usage, these templates are restricted to certain wavelengths.
In the near-infrared (NIR) to ultraviolet (UV) wavelengths (the region where the energy output of stars peaks), the spectrum\footnote{In the remainder of this paper, we are concerned with a limited range in wavelength, and so we refer to spectra rather than spectral energy distributions.} contains information about the main physical properties of galaxies, e.g., age, star formation rate (SFR), stellar mass, metallicity, and interstellar medium. 
\textbf{
Using templates to classify galaxy spectra can help to answer the question of whether galaxy spectra form a continuous distribution or can  be separated into discrete groups.
Templates are also helpful in computing $K$-corrections and discovering new classes of objects as outliers \citep{Folkes96}.
}

@@

A Kohonen self-organizing map (also called self-organizing map, or SOM) is a (semi)-supervised neural network for mapping and visualizing a complex and nonlinear high dimension data introduced by~\citet{Kohonen82}.
It shows simple geometrical relationships in non-linear high dimensional data on a map \citep{Kohonen98}.
\textbf{The training of a self-organizing map is fully unsupervised.
     Using the resulting map as a template to classify other data requires labelling the output, which is why some groups consider self-organizing maps to be a (semi)-supervised method.}

@@

\textbf{
The goal of this work is to investigate the use of self-organizing maps in classifying galaxy spectra against templates, and compare them to other methods. 
The questions we want to answer are: 
(1) What is the degree of agreement between $\chi^2$ fitting, one- and two-dimensional self-organizing maps, K-means clustering and artificial neural networks? 
(2) Are certain spectral types more (or less) consistently classified by the different methods? 
(3) To what extent do the average properties of galaxy classes represent a continuous distribution of physical properties?
We analyze the same sample of 142 galaxies with $0.5 < z < 1$ as 
\citet[][hereafter \citetalias{Hossein12}]{Hossein12}, who classified the spectra using a supervised (multi-layer feed-forward) neural network method with the \citetalias{Kinney96} templates.
For each of the 142 galaxies, we find the best chi-squared fit among the 12 templates  to have a base for comparing the methods' performance.
We also use the Fleiss Kappa index \citet{Fleiss71} to test interrater reliability 
between the three methods.
% HAVE NOT YET DONE THE FOLLOWING. MAY WANT TO INCLUDE COMMENTED SENTENCES.
We calculate the dispersion in properties between galaxies classified in the same groups in order to measure how well each classification method groups similar objects together.
}

% As in \citetalias{Hossein12}, we examine the averaged properties of the new galaxy groups and compare them with previous work.
%\citetalias{Hossein12} also 
%showed that there are tight correlations between physical properties of galaxies, and these %correlations might be different for each type of galaxy.

>>>>

Several attempts have been made to create a complete set of templates for categorizing the spectral type of galaxies using observations of nearby galaxies~(e.g.~\citealt{Kinney93}, \citealt[][hereafter \citetalias{Kinney96}]{Kinney96}, \citealt{Bershady00}, \citealt{Mannucci01}). 
Based on their usage, these templates are restricted to certain wavelengths.
In the near-infrared (NIR) to ultraviolet (UV) wavelengths (the region where the energy output of stars peaks), the spectrum\footnote{In the remainder of this paper, we are concerned with a limited range in wavelength, and so we refer to spectra rather than spectral energy distributions.} contains information about the main physical properties of galaxies, e.g., age, star formation rate (SFR), stellar mass, metallicity, and interstellar medium.

Using templates to classify galaxy spectra can help to answer the question of whether galaxy spectra form a continuous distribution or can  be separated into discrete groups.
Templates are also helpful in computing $K$-corrections and discovering new classes of objects as outliers \citep{Folkes96}.

@@

A Kohonen self-organizing map (also called self-organizing map, or SOM) is a (semi)-supervised neural network for mapping and visualizing a complex and nonlinear high dimension data introduced by~\citet{Kohonen82}.
It shows simple geometrical relationships in non-linear high dimensional data on a map \citep{Kohonen98}.
The training of a self-organizing map is fully unsupervised.
     Using the resulting map as a template to classify other data requires labelling the output, which is why some groups consider self-organizing maps to be a (semi)-supervised method.

@@

The goal of this work is to investigate the use of self-organizing maps in classifying galaxy spectra against templates, and compare them to other methods. 
The questions we want to answer are: 
(1) What is the degree of agreement between classifications produced $\chi^2$ fitting, one- and two-dimensional self-organizing maps, K-means clustering and artificial neural networks? 
(2) Are certain spectral types more (or less) consistently classified by the different methods? 
(3) Which classification methods produce classes with the highest internal similarity ?
We analyze the same sample of 142 galaxies with $0.5 < z < 1$ as 
\citet[][hereafter \citetalias{Hossein12}]{Hossein12}, who classified the spectra using a supervised (multi-layer feed-forward) neural network method with the \citetalias{Kinney96} templates.
For each of the 142 galaxies, we find the best chi-squared fit among the 12 templates to use as a base for comparing the methods' performance.
We compute an agreement score for each method and also use the Fleiss Kappa index \citet{Fleiss71} to test interrater reliability 
between the methods.
To measure how well each classification method groups similar objects together, we
calculate the silhouette scores and dispersions in properties between galaxies classified in the same groups.

### 2017-08-11 sections0.08_for_editor/method0.08.tex
re-submitted version

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Method
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

@@

\textbf{In this section we describe the algorithms for both self-organizing maps and K-means clustering. As K-means is both simpler and more well-known, we introduce it briefly first before discussing SOM in detail.}

@@

\textbf{
The K-means method partitions a data set $x_i$, having $n$ points in $d$-dimensional feature space, into $K$ clusters.
The number of clusters $K<n$ must be decided by the user in advance, and the cluster centroids in feature space are located at positions $\mu_k$.
The goal of the algorithm is to minimize the sum
\begin{equation}
\label{eq:km_sos}
J = \sum_{i=1}^n \sum_{k=1}^K \min \big( \norm{x_i - \mu_k}^2 \big)
\end{equation}
where $\norm{z}$ indicates the distance measurement in $d$-dimensional space. In this work the Euclidean distance is is used, but other metrics are also possible.
The algorithm is initialized by randomly choosing $K$ random points as cluster centroids.
Each data point is assigned to the cluster with the closest centroid, and the centroids are re-calculated.
The steps of re-assigning cluster membership and re-calculating the cluster centroids are repeated until a stopping criterion is reached (a maximum number of iterations or no change in cluster assignments).
We used the \textsc{matlab} K-means library~\citep{Seber84, Spath85}, written for \textsc{matlab2015b}, to perform  K-means clustering.
The default maximum number of iterations (100) was used.
}

@@

     We start by creating S1 $\times$ S2 empty neurons. 
     The arrangement of these neurons depends on the map's topology provided by the user. 
     In the case of 1D maps, since each neuron has two immediate neighbours, the topology of the map does not have any effect on the final result and any topology can be chosen.
     However, in 2D maps, the shape of the neurons specifies the number of immediate neighbours for each neuron and it is up to user to choose the most suitable shape based on the data.
     In this paper, we choose hexagonal topology, which gives each neuron six neighbours, and provides more interactions between neurons.
     Initially a random weight vector, \boldit{W} $\in \Re^d$, will be assigned to each node.
     The process of creating SOM happens over a series of $N$ iterations, where $N$ is set by the user. 
     During each iteration the weight vectors might change according to the Kohonen learning rule (equation~\ref{equ: weight adj}). 
      In each iteration, the SOM code:
     \begin{enumerate}
        \item chooses a random vector from the dataset ($V_i$).
        \item calculates the Euclidean distance in $\Re^d$ space for each node, $j$, as  $D_j^2= \sum_{i=0}^{i=d} (V_i - W_{j_i})^2$, and finds a neuron with minimum $D_j$, (``$D_{j_{\min}}$"). This neuron is the winner node and is called the Best Matching Unit (BMU). 
        \item  computes the radius of the neighbourhood of the BMU to find nodes within this radius. The weight vectors of these nodes will be affected in the next steps. The radius of the neighbourhood is arbitrary and can be set to be as high as half of the SOM size. It then decays exponentially over each iteration as
        \begin{equation}
            r^t_{\rm BMU} = r^0_{\rm BMU}e^{(-t/\tau)}
        \end{equation}
        where $\tau$ is a decay constant and is usually set to be the same as the number of iterations, $N$. $r^0_{BMU}$ and $r^t_{\rm BMU}$ are the radii of the neighbourhood at 0th and $t$th iteration, respectively. 
        \item changes the weight vectors of the BMU and all the nodes within $r^t_{\rm BMU}$ as:
        %Sahar: should w and v in the equation below be W and V (to match the text)?
        \begin{equation}
            \label{equ: weight adj}
            w(t+1)=w(t)+L(t) \times R(t) \times(v(t)-w(t))
        \end{equation}
        where $L(t) = L_0 e^{(-t/\tau)}$ is the learning factor, which prevents the divergence of the SOM and $R(t)=\exp(-\frac{D_j^2}{2r^t_{\rm BMU}})$ is the influence rate. $R(t)$ determines how the weight of each node in the neighbourhood of BMU will change.
     \end{enumerate}
     These steps are then repeated $N$ times.

@@

     \textbf{As mentioned in the introduction, one of the main advantages of the SOM method is that the resulting networks can be used to cluster new datasets with no additional training required.
     New data with the same dimensionality as the original input data can be presented to the already trained network: the SOM algorithm finds the best matching unit considering the weight of nodes in the trained network and each vector from the new data set.
     The winner node determines the place of each new vector on the SOM.
     The location of the new vectors on the SOM allows them to be compared with the original data set. }

>>>>

@@

In this section we describe the algorithms for both self-organizing maps and K-means clustering. As K-means is both simpler and more well-known, we introduce it briefly first before discussing SOM in detail.

@@

The K-means method partitions a data set $x_i$, having $n$ points in $d$-dimensional feature space, into $K$ clusters.
The number of clusters $K<n$ must be decided by the user in advance, and the cluster centroids in feature space are located at positions $\mu_k$.
The goal of the algorithm is to minimize the sum
\begin{equation}
\label{eq:km_sos}
J = \sum_{i=1}^n \sum_{k=1}^K \min \big( \norm{x_i - \mu_k}^2 \big)
\end{equation}
where $\norm{z}$ indicates the distance measurement in $d$-dimensional space. In this work the Euclidean distance is is used, but other metrics are also possible.
The algorithm is initialized by randomly choosing $K$ random points as cluster centroids.
Each data point is assigned to the cluster with the closest centroid, and the centroids are re-calculated.
The steps of re-assigning cluster membership and re-calculating the cluster centroids are repeated until a stopping criterion is reached (a maximum number of iterations or no change in cluster assignments).
We used the \textsc{matlab} K-means library~\citep{Seber84, Spath85}, written for \textsc{matlab2015b}, to perform  K-means clustering.
The default maximum number of iterations (100) was used.

@@

     We start by creating S1 $\times$ S2 empty neurons. 
     The arrangement of these neurons depends on the map's topology provided by the user. 
     In the case of 1D maps, since each neuron has two immediate neighbours, the topology of the map does not have any effect on the final result and any topology can be chosen.
     However, in 2D maps, the shape of the neurons specifies the number of immediate neighbours for each neuron and it is up to user to choose the most suitable shape based on the data.
     In this paper, we choose hexagonal topology, which gives each neuron six neighbours, and provides more interactions between neurons.
     Initially a random weight vector, \boldit{W} $\in \Re^d$, will be assigned to each node.
     The process of creating SOM happens over a series of $N$ iterations, where $N$ is set by the user. 
     During each iteration the weight vectors might change according to the Kohonen learning rule (equation~\ref{equ: weight adj}). 
      In each iteration, the SOM code:
     \begin{enumerate}
        \item chooses a random vector from the dataset ($V_i$).
        \item calculates the Euclidean distance in $\Re^d$ space for each node, $j$, as  $D_j^2= \sum_{i=0}^{i=d} (V_i - W_{j_i})^2$, and finds a neuron with minimum $D_j$, (``$D_{j_{\min}}$"). This neuron is the winner node and is called the Best Matching Unit (BMU). 
        \item  computes the radius of the neighbourhood of the BMU to find nodes within this radius. The weight vectors of these nodes will be affected in the next steps. The radius of the neighbourhood is arbitrary and can be set to be as high as half of the SOM size. It then decays exponentially over each iteration as
        \begin{equation}
            r^t_{\rm BMU} = r^0_{\rm BMU}e^{(-t/\tau)}
        \end{equation}
        where $\tau$ is a decay constant and is usually set to be the same as the number of iterations, $N$. $r^0_{BMU}$ and $r^t_{\rm BMU}$ are the radii of the neighbourhood at 0th and $t$th iteration, respectively. 
        \item changes the weight vectors of the BMU and all the nodes within $r^t_{\rm BMU}$ as:
        \begin{equation}
            \label{equ: weight adj}
            w(t+1)=w(t)+L(t) \times R(t) \times(v(t)-w(t))
        \end{equation}
        where $L(t) = L_0 e^{(-t/\tau)}$ is the learning factor, which prevents the divergence of the SOM and $R(t)=\exp(-\frac{D_j^2}{2r^t_{\rm BMU}})$ is the influence rate. $R(t)$ determines how the weight of each node in the neighbourhood of BMU will change.
     \end{enumerate}
     These steps are then repeated $N$ times.

@@

     As mentioned in the introduction, one of the main advantages of the SOM method is that the resulting networks can be used to cluster new datasets with no additional training required.
     New data with the same dimensionality as the original input data can be presented to the already trained network: the SOM algorithm finds the best matching unit considering the weight of nodes in the trained network and each vector from the new data set.
     The winner node determines the place of each new vector on the SOM.
     The location of the new vectors on the SOM allows them to be compared with the original data set.

     \subsection{Classification and clustering metrics}
     \label{sec:metrics}

     We use several metrics to measure the degree of agreement between classification methods. 
     As a baseline method for assigning galaxy spectra to template classes, we compute the value of chi-squared between each galaxy and each template and assign the galaxy to the template class with lowest chi-squared value. 
     A metric of agreement with this classification is an agreement score, computed as follows.
    If a method sorted a galaxy into the same cluster or neuron as its best-fit chi-square template, this was considered a match and scored 1 point. 
    For the larger-sized self-organized maps, if a galaxy was classified between two nodes, one of which contained its best-fit chi-square template, this scored 0.5 point. 
    Otherwise, the classification score was zero points.     
     Another method to check the degree of agreement between different methods is the Fleiss kappa index \citet{landis77}. 
    This index measures agreement between three or more raters that are classifying a specific number of objects to a number of classes and was calculated using the R software package.

        Classification and clustering are related techniques, and we make
        another comparison between the classification methods with a clustering metric, the silhouette score  \citep{rousseeuw87}.        
This score is a metric used to describe cluster compactness and isolation, given by:
\begin{equation}
S = \frac{1}{n} \sum{\frac{b - a}{\max\big(a, b\big)}}
\end{equation}
where $a$ is the mean distance between a point and the other points in its cluster, and $b$ is the mean distance between a point and the nearest cluster of which that point is not a member.
A higher silhouette score corresponds to a partition with better defined clusters.
For a given dataset, the score generally declines with the number of clusters and is sometime used to determine the optimal cluster number.
In this work we use the silhouette score to compare classifications with a comparable number of classes, as a measure of similarity within the classes.
Mean values for each classification were calculated using the \textsc{scikit-learn} Python package \citep{sklearn}.
